---
title: | 
  | Supplemental Material for 
  | 'Niche estimation above and below the species level'

author:  

- name: Adam B. Smith
  affilnum: 1
  email: adam.smith@mobot.org

- name: William Godsoe
  affilnum: 2

- name: Francisco Rodríguez-Sánchez 
  affilnum: 3

- name: Hsiao-Hsuan Wang
  affilnum: 4

- name: Dan Warren
  affilnum: '5,6'
  

affiliation:
  
- affilnum: 1
  affil: Center for Conservation and Sustainable Development, Missouri Botanical Garden, 4344 Shaw Boulevard, Saint Louis, MO 63116 USA
  
- affilnum: 2
  affil: BioProtection Research Centre, Burns Building Lincoln University, Ellesmere Junction Road/Springs Road Lincoln University, Lincoln, New Zealand
  
- affilnum: 3
  affil: Department of Integrative Ecology, Estación Biológica de Doñana, Consejo Superior de Investigaciones Científicas, Avda. Américo Vespucio 26, 41092 Sevilla, Spain
  
- affilnum: 4
  affil: Department of Wildlife and Fisheries Sciences, Texas A&M University, 534 John Kimbrough Blvd., Building 1537, 2258 TAMU, College Station, Texas 77843, USA
  
- affilnum: 5
  affil: Senckenberg Biodiversity and Climate Research Center (SBiK-F), Frankfurt am Main, Germany
  
- affilnum: 6
  affil: Biodiversity and Biocomplexity Unit, Okinawa Institute of Science and Technology, Japan


output:

  pdf_document:
    fig_caption: yes
    fig_width: 5
    fig_height: 5
    keep_tex: yes
    number_sections: no
    template: manuscript.latex
    includes:
      in_header: header.tex

documentclass: article
classoption: a4paper
capsize: normalsize
fontsize: 11pt
geometry: margin=1in
linenumbers: no
spacing: doublespacing
footerdate: no
abstract: no
---


## R code for comparing splitting, lumping, partial pooling, and phylogenetically-informed ecological niche models on a simulated dataset (*Box 3*)

This source code is archived in Zenodo (Rodríguez-Sánchez 2018) and can also be found at https://github.com/Pakillo/phyloENM-TREE.


### Simulate taxa niches and phylogeny

Here we simulate (following Pearse *et al.* 2016) a group of phylogenetically related taxa with different sentivities to temperature, so that closely related taxa will have more similar niches.


```{r}

nspp <- 6
nsite <- 30

# simulate a phylogenetic tree
library(ape)
set.seed(6)
phy <- rtree(n = nspp)
phy <- compute.brlen(phy, method = "Grafen", power = 0.5)
#plot(phy)

# standardize the phylogenetic covariance matrix to have determinant 1
Vphy <- vcv(phy)
Vphy.std <- Vphy/(det(Vphy)^(1/nspp))

# Perform a Cholesky decomposition of Vphy 
iD <- t(chol(Vphy.std))

# Generate environmental site variable
env <- seq(0, 20, length.out = nsite)


## Set up species-specific regression coefficients as random effects

set.seed(79)

intercept <- iD %*% runif(nspp, -1, 2)
slope <- iD %*% runif(nspp, -0.4, 0.1)

intercept <- intercept[gtools::mixedorder(rownames(intercept)), ]
slope <- slope[gtools::mixedorder(rownames(slope)), ]

#intercept
#slope

## Calculate suitabilities for each taxa and site
suitab <- rep(intercept, each = nsite)
suitab <- suitab + rep(slope, each = nsite) * rep(env, nspp)
suitab.error <- suitab + rnorm(nspp * nsite, mean = 0, sd = 1) #add some random 'error'  
suitab.invlogit <- arm::invlogit(suitab)
suitab.error.invlogit <- arm::invlogit(suitab.error)

set.seed(114)

pres <- rbinom(length(suitab.error), size = 1, prob = suitab.error.invlogit)  # pres-abs

dat <- data.frame(taxa = paste("t", sort(rep(1:nspp, nsite)), sep = ""),
                  site = rep(1:nsite, nspp),
                  env = rep(env, nspp), 
                  suitab.invlogit = suitab.invlogit,
                  presabs = pres)

```


### Plot simulated niches

```{r message = FALSE}

library(ggplot2)
library(cowplot)
library(viridis)


simul.niches <- ggplot(dat, aes(env, suitab.invlogit, colour = taxa)) +
  ylim(0, 1) +
  labs(x = "Temperature", y = "Suitability", title = "True niches") +
  geom_line(size = 2) +
  scale_color_viridis(discrete = TRUE) + 
  theme(plot.title = element_text(size = 18)) +
  theme(legend.position = "none")

simul.niches

```


### Plot phylogeny

```{r message=FALSE}

library(ggtree)

phylog <- ggplot(phy) + 
  geom_tree(size = 2, colour = "grey70") +
  geom_tree(size = 2, aes(color = label)) + 
  scale_color_viridis(discrete = "TRUE") +
  theme(legend.position = "none",
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank()) +
  labs(title = "Phylogeny") +
  theme(plot.title = element_text(size = 18)) +
  theme(plot.margin = unit(c(0.2,1,1,1), "cm"))

phylog

```


### Splitting

Here we fit independent binomial generalized linear models (GLM) to each taxon.

```{r}

split <- ggplot(dat, aes(env, presabs, colour = taxa)) +
  ylim(0, 1) +
  labs(x = "Temperature", y = "Suitability", title = "Splitting (GLM)") +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE, size = 2) +
  scale_color_viridis(discrete = TRUE) +
  theme(plot.title = element_text(size = 18)) +
  theme(legend.position = "none")

split

```


### Lumping

Here we fit a single binomial generalized linear model (GLM) to all taxa together.

```{r}

lump <- ggplot(dat, aes(env, presabs)) +
  ylim(0, 1) +
  labs(x = "Temperature", y = "Suitability", title = "Lumping (GLM)") +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE, size = 3,
              colour = "black") +
  # for comparison with real niches:
  geom_line(aes(env, suitab.invlogit, colour = taxa), size = 2, alpha = 0.2) +
  scale_color_viridis(discrete = TRUE, guide = guide_legend(override.aes = list(alpha = 1))) +
  theme(plot.title = element_text(size = 18)) +
  theme(legend.position = "none")

lump

```
  

### Pooling (GLMM)

Here we fit a mixed effects model (GLMM) with varying intercepts and slopes, so that there is partial pooling among taxa. 

```{r message=FALSE}

library(lme4)

mixed <- glmer(presabs ~ env + (1 + env | taxa), data = dat, family = binomial)
# summary(mixed)
# coef(mixed)
dat$mixedpred <- fitted(mixed, newdata = dat, type = "response")

mixedfig <- ggplot(dat, aes(env, mixedpred, colour = taxa)) +
  ylim(0, 1) +
  labs(x = "Temperature", y = "Suitability", title = "Pooling (GLMM)") + 
  geom_line(size = 2) +
  scale_color_viridis(discrete = TRUE) +
  theme(plot.title = element_text(size = 18)) +
  theme(legend.position = "none")

mixedfig

```



### Phylogenetic pooling (PGLMM)

Here we fit a phylogenetic generalized linear model (PGLMM) with varying intercepts and slopes, so that closely related taxa tend to have more similar niches. 

```{r cache=TRUE, message=FALSE}

inv.phylo <- MCMCglmm::inverseA(phy, nodes = "TIPS", scale = TRUE)  
A <- solve(inv.phylo$Ainv)
rownames(A) <- rownames(inv.phylo$Ainv)

library(brms)

pglmm <- brm(
  presabs ~ env + (1 + env | taxa), 
  data = dat, 
  family = bernoulli(), 
  cov_ranef = list(taxa = A),  
  prior = c(
    prior(normal(0, 3), "b"),
    prior(normal(0, 3), "Intercept"),
    prior(student_t(3, 0, 3), "sd")),  # sd of group random effect
  chains = 3, cores = 3, iter = 2000,
  control = list(adapt_delta = 0.96)
)


# summary(pglmm)
# plot(pglmm)
# coef(pglmm)
# pp_check(pglmm)

dat$pglmm <- fitted(pglmm, newdata = dat, scale = "response")[,1]
```

```{r}

pglmmfig <- ggplot(dat, aes(env, pglmm, colour = taxa)) +
  ylim(0, 1) +
  labs(x = "Temperature", y = "Suitability", title = "Phylogenetic Pooling (PGLMM)") + 
  geom_line(size = 2) +
  scale_color_viridis(discrete = TRUE) +
  theme(plot.title = element_text(size = 18)) +
  theme(legend.position = "none")

pglmmfig

```


### Combined figure

```{r out.height='9in', out.width='7in'}

figure <- plot_grid(simul.niches, phylog, 
                    split, lump,
                    mixedfig, pglmmfig, 
                    nrow = 3, ncol = 2)

#figure 

save_plot(paste0("FigureBox3", ".pdf"), figure, base_height = 11, base_width = 9)

```



### References and software used


Bates, Douglas, Martin Mächler, Ben Bolker, and Steve Walker. 2015. “Fitting Linear Mixed-Effects Models Using lme4.” Journal of Statistical Software 67 (1): 1–48. doi:10.18637/jss.v067.i01.

Bürkner, Paul-Christian. 2017. “brms: An R Package for Bayesian Multilevel Models Using Stan.” Journal of Statistical Software 80 (1): 1–28. doi:10.18637/jss.v080.i01.

Garnier, Simon. 2018. Viridis: Default Color Maps from ’Matplotlib’. https://CRAN.R-project.org/package=viridis.

Hadfield, Jarrod D. 2010. “MCMC Methods for Multi-Response Generalized Linear Mixed Models: The MCMCglmm R Package.” Journal of Statistical Software 33 (2): 1–22. http://www.jstatsoft.org/v33/i02/.

Paradis, E., J. Claude, and K. Strimmer. 2004. “APE: Analyses of Phylogenetics and Evolution in R Language.” Bioinformatics 20: 289–90.

Pearse WD, MW Cadotte, J Cavender-Bares, AR Ives, C Tucker, S Walker & MR Helmus. 2016. A gentle introduction to Phylogenetic Generalised Linear Mixed Models. https://cran.r-project.org/web/packages/pez/vignettes/pez-pglmm-overview.pdf

R Core Team. 2018. R: A Language and Environment for Statistical Computing. Vienna, Austria: R Foundation for Statistical Computing. https://www.R-project.org/.

Rodríguez-Sánchez F. 2018. R code for comparing splitting, lumping, partial pooling, and phylogenetically-informed ecological niche models (version 1.1). Zenodo. http://doi.org/10.5281/zenodo.1470855 

Wickham, Hadley. 2009. Ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag New York. http://ggplot2.org.

Wilke, Claus O. 2017. Cowplot: Streamlined Plot Theme and Plot Annotations for ’Ggplot2’. https://CRAN.R-project.org/package=cowplot.

Xie, Yihui. 2018. Knitr: A General-Purpose Package for Dynamic Report Generation in R. https://yihui.name/knitr/.

Yu, Guangchuang, David Smith, Huachen Zhu, Yi Guan, and Tommy Tsan-Yuk Lam. 2017. “Ggtree: An R Package for Visualization and Annotation of Phylogenetic Trees with Their Covariates and Other Associated Data.” Methods in Ecology and Evolution. doi:10.1111/2041-210X.12628.






### Details on computation environment

```{r echo=FALSE}
sessioninfo::session_info()
```

